---
title: "Hypothesis testing"
format: revealjs
---

```{r}
#| label: setup
#| include: false
ggplot2::theme_set(ggplot2::theme_gray(base_size = 16))
```

# Setup

## Packages

- **tidyverse** for data wrangling and visualization
- **tidymodels** for modeling
- **openintro** for data

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(tidymodels)
library(openintro)
```

# Hypothesis testing

## Hypothesis testing

A hypothesis test is a statistical technique used to evaluate *competing claims* using data

::: incremental
-   **Null hypothesis,** $H_0$: An assumption about the population.
    "There is nothing going on."

-   **Alternative hypothesis,** $H_A$: A research question about the population.
    "There is something going on".
:::

. . .

Note: Hypotheses are always at the population level!

## Setting hypotheses

-   **Null hypothesis,** $H_0$: "There is nothing going on." The slope of the model for predicting the prices of houses in Duke Forest from their areas is 0, $\beta_1 = 0$.

-   **Alternative hypothesis,** $H_A$: "There is something going on".
    The slope of the model for predicting the prices of houses in Duke Forest from their areas is different than, $\beta_1 \ne 0$.

## Hypothesis testing "mindset"

-   Assume you live in a world where null hypothesis is true: $\beta_1 = 0$.

-   Ask yourself how likely you are to observe the sample statistic, or something even more extreme, in this world: $P(b_1 \leq -159~or~b_1 \geq 159 | \beta_1 = 0)$ = ?

## Hypothesis testing as a court trial

-   **Null hypothesis**, $H_0$: Defendant is innocent

-   **Alternative hypothesis**, $H_A$: Defendant is guilty

. . .

-   **Present the evidence:** Collect data

. . .

-   **Judge the evidence:** "Could these data plausibly have happened by chance if the null hypothesis were true?"
    -   Yes: Fail to reject $H_0$
    -   No: Reject $H_0$

## Hypothesis testing framework

::: incremental
-   Start with a null hypothesis, $H_0$, that represents the status quo

-   Set an alternative hypothesis, $H_A$, that represents the research question, i.e. what we’re testing for

-   Conduct a hypothesis test under the assumption that the null hypothesis is true and calculate a **p-value** (probability of observed or more extreme outcome given that the null hypothesis is true)

    -   if the test results suggest that the data do not provide convincing evidence for the alternative hypothesis, stick with the null hypothesis
    -   if they do, then reject the null hypothesis in favor of the alternative
:::

## Calculate observed slope

...
which we have already done:

```{r}
observed_fit <- duke_forest |>
  specify(price ~ area) |>
  fit()

observed_fit
```

## Simulate null distribution

```{r}
#| code-line-numbers: "|1|2|3|4|5|6"
set.seed(12345)
null_dist <- duke_forest |>
  specify(price ~ area) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  fit()
```

## View null distribution {.smaller}

```{r}
null_dist
```

## Visualize null distribution {.smaller}

```{r}
null_dist |>
  filter(term == "area") |>
  ggplot(aes(x = estimate)) +
  geom_histogram(binwidth = 15)
```

## Visualize null distribution (alternative)

```{r}
visualize(null_dist) +
  shade_p_value(obs_stat = observed_fit, direction = "two-sided")
```

## Get p-value {.smaller}

```{r}
null_dist |>
  get_p_value(obs_stat = observed_fit, direction = "two-sided")
```

## Make a decision

::: task
Based on the p-value calculated, what is the conclusion of the hypothesis test?
:::
