---
title: Logistic regression
subtitle: "{{< fa book-open >}} Modeling Inference <br>{{< fa book >}} Data Science with R"
format: live-html
engine: knitr
bibliography: references.bib
webr:
  packages: 
    - tidyverse
    - tidymodels
    # - faraway
    #- scales
    #- ggridges
    # - kableExtra
  cell-options:
    autorun: false
  resources:
    - https://raw.githubusercontent.com/data-science-with-r/modeling-inference/refs/heads/main/exercises/data/income.csv
---

<!-- begin: webr fodder -->

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| edit: false
#| echo: false
#| output: false
options("readr.edition" = 1)
library(tidyverse)
# Read the CSV file directly from the URL
income <- read_csv("https://raw.githubusercontent.com/data-science-with-r/modeling-inference/refs/heads/main/exercises/data/income.csv")
```

<!-- end: webr fodder -->

# Getting Started

Programming exercises are designed to provide an opportunity for you to put what you learn in the videos and readings. These exercises feature interactive code cells which allow you to write, edit, and run R code without leaving your browser.

When the ▶️ Run Code button turns to a solid color (with no flashing bubble indicating that the document is still loading), you can interact with the code cells!

## Packages

We’ll use the **tidyverse** and **tidymodels** for this programming exercise. These are already installed for you to use!

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
# library(faraway)
library(tidymodels)
library(kableExtra)
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

income <- read_csv("data/income.csv") 
```

## Motivation

Census data can help us predict income based on other variables, such as demographic, social, and economic variables. We can use this information to help gain deeper insights into the economic well-being of a population, which can help practically inform items such as strategic planning, resource allocation, and market analysis. In this programming exercise, we are going to explore the relationship between different predictors, and if an individual was making more or less than \$50,000 a year in 1994 in the United States.

## Dataset information

The dataset we will be using for this exercise is from the 1994 Census database, extracted by Barry Becker. These data can be found on the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/2/adult). Expand the tab below to look at the data key.

```{webr}
#| label: load-data
income <- read_csv("income.csv") 

income
```


::: callout-caution
|                |                                                                                                                                                                                                                                                                                                                                                                                                                                 |
|------------------|------------------------------------------------------|
| Variable Name  | Description                                                                                                                                                                                                                                                                                                                                                                                                                     |
| age            | Age                                                                                                                                                                                                                                                                                                                                                                                                                             |
| workclass      | Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, State-gov, Without-pay, Never-worked.                                                                                                                                                                                                                                                                                                                          |
| education      | Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, Assoc-voc, 9th, 7th-8th, 12th, Masters, 1st-4th, 10th, Doctorate, 5th-6th, Preschool.                                                                                                                                                                                                                                                                          |
| education-num  | Education Level as a number                                                                                                                                                                                                                                                                                                                                                                                                     |
| marital-status | Married-civ-spouse, Divorced, Never-married, Separated, Widowed, Married-spouse-absent, Married-AF-spouse.                                                                                                                                                                                                                                                                                                                      |
| occupation     | Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, Prof-specialty, Handlers-cleaners, Machine-op-inspct, Adm-clerical, Farming-fishing, Transport-moving, Priv-house-serv, Protective-serv, Armed-Forces.                                                                                                                                                                                                       |
| relationship   | Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.                                                                                                                                                                                                                                                                                                                                                             |
| race           | White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.                                                                                                                                                                                                                                                                                                                                                                    |
| sex            | Female, Male.                                                                                                                                                                                                                                                                                                                                                                                                                   |
| native-country | United-States, Cambodia, England, Puerto-Rico, Canada, Germany, Outlying-US(Guam-USVI-etc), India, Japan, Greece, South, China, Cuba, Iran, Honduras, Philippines, Italy, Poland, Jamaica, Vietnam, Mexico, Portugal, Ireland, France, Dominican-Republic, Laos, Ecuador, Taiwan, Haiti, Columbia, Hungary, Guatemala, Nicaragua, Scotland, Thailand, Yugoslavia, El-Salvador, Trinadad&Tobago, Peru, Hong, Holand-Netherlands. |
| income         | \>50K, \<=50K.                                                                                                                                           :::

## Exploratory Data Analysis

::: exercise

Before we fit a logistic regression model, we are going to explore these data. Let' start with exploring our response variable of interest, cholesterol (`income`). For this exercise, we are going to investigate the relationship between `income` and `age`. 

Let's calculate the mean and sd of `age` for each level of `income`. What does this tell us? When fitting a logistic regression model, would we expect the probability of making more than \$50,000 to increase or decrease as you get older?

```{webr}

```
:::

::: {.callout-tip collapse="true"}
## Solution

```{r}
income |>
  group_by(income) |>
  summarize(mean_age = mean(age),
            sd_age = sd(age))
```

The mean age for those making less than \$50,000 is roughly 7.4 years younger than the mean age of those making more than \$50,000. The spread (standard deviation) is also higher for those making less than \$50,000 in comparison to those making more. 

Based on this information, we would expect the probability of making more than \$50,000 to increase as age increases. 
:::

### Fitting a logistic regression model
::: exercise
Fit a logisitc regression model below with income as our response, and age as our predictor. Compare your output to the written out model below. Is your respnose the log-odds of making more than \$50,000, or the probability of making more that \$50,000?

:::

```{webr}

```



$$
\log \left( \frac{p}{1-p} \right) = -2.74 + 0.0395(\text{age})
$$

::: {.callout-tip collapse="true"}
## Solution
```{r}
income_model <- logistic_reg() |>
  set_engine("glm") |>
  fit(income ~ age , data = income, family = binomial)

tidy(income_model)
```

The logistic regression model we fit models the log-odds: $$\log \left( \frac{p}{1-p} \right)$$, where `p` represents the probability of of making more than \$50,000. 
:::

#### Probabilities 

To model the probability of making more than \$50,000, we need to expoentiate both sides of the eqaution. See below: 

$$
p = \frac{e^{-2.74 + 0.0395(\text{age})}}{1 + e^{-2.74 + 0.0395(\text{age})}}
$$

::: exercise
Using your model, what's the probability of making more than \$50,000 if you are 60 years old in 1994?

```{webr}

```
:::

::: {.callout-tip collapse="true"}
## Solution

$$
p = \frac{e^{-2.74 + 0.0395(\text{60})}}{1 + e^{-2.74 + 0.0395(\text{60})}} = 0.408
$$

We can also find this in R! 

```{webr}
age_predict <- tibble(age = 60)

predict(income_model$fit, age_predict, type = "response") 
```
:::

Based on your concoulsion from your exploratory analysis, does the probabilty of making over \$50,000 increase or decrease as you age? How can you tell from your model?

::: {.callout-tip collapse="true"}
## Solution
The coefficient assocciated with age is positive, meaning that the estimated probability of making more than \$50,000 increases as age increases. 
:::

We can further show your intuition is correct by calculating the probability at the `age` of 61, and see if the difference between the probability at the age of 60 is positive or negative. 

$$
p = \frac{e^{-2.74 + 0.0395(\text{61})}}{1 + e^{-2.74 + 0.0395(\text{61})}} = 0.418

.418 - .408 = .1
$$

The estimated probabilty of making more than \$50,000 increases by roughly 1 percent as you go from age 60 to 61.

```{r}
#| echo: false
# Define the logistic function based on your model's coefficients
calculate_probability <- function(age) {
  intercept <- -2.74
  beta_age <- 0.0395
  exponent <- intercept + beta_age * age
  probability <- exp(exponent) / (1 + exp(exponent))
  return(probability)
}

# Create a data frame with a range of ages
plot_data <- tibble(age = seq(18, 90, by = 0.5)) |>
  mutate(probability = calculate_probability(age))

# Plot the model's predicted probabilities
ggplot(plot_data, aes(x = age, y = probability)) +
  geom_line(color = "dodgerblue", linewidth = 1.2) +
  labs(
    title = "Predicted Probability of >$50k by Age",
    x = "Age",
    y = "Probability of >$50k "
  ) +
  theme_minimal() +
  ylim(0, 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray") +
  annotate("text", x = 80, y = 0.52, label = "", vjust = -0.5, color = "gray")
```

### Your Turn 

There are many other factors that could contribute to the probability of making more than \$50,000 than just age in 1994. Now, we challenge you to fit a more complicated model to help better understand which factors make help best explain our response variable.
